\chapter{Appendix}
\label{chap:appendix}

\section{The Termination Tradeoff}
\label{sec:termination-tradeoff}

This appendix formalizes the core semantic tradeoff underlying consensusless protocols: replacing the unconditional termination guarantee of classical Byzantine consensus with a weaker, conflict-conditioned variant. We first recall the standard definition of single-shot consensus and then present the relaxed formulation adopted by protocols such as ABC~\cite{abc} and FastPay~\cite{fastpay}, highlighting how this relaxation enables the performance gains discussed throughout the thesis while preserving safety unconditionally.

We start from the classical definition of single-shot consensus.

\textbf{Single-shot consensus.}
In a single-shot consensus instance, parties must decide on a single value (e.g., one proposal among potentially conflicting candidates).

\begin{definition}[Single-shot Consensus]
A protocol solves Byzantine consensus if it satisfies:
\begin{itemize}
    \item \textbf{Agreement.} No two honest parties decide different values.
    \item \textbf{Validity.} If all honest parties start with the same value $v$, then $v$ is the only value that can be decided by honest parties.
    \item \textbf{Termination.} Every honest party eventually decides some value.
\end{itemize}
\end{definition}

Termination is unconditional: progress must occur even under conflicts. This coordination requirement is a primary contributor to the latency and throughput costs of consensus-backed blockchains.

\textbf{Relaxed termination.}
Consensusless protocols improve performance by weakening termination: instead of requiring progress under arbitrary conflicts, they require progress only when there is no conflict~\cite{abc}.

\begin{definition}[Consensus with Relaxed Termination~\cite{abc}]
A protocol satisfies relaxed consensus if it provides:
\begin{itemize}
    \item \textbf{Agreement.} No two honest parties decide different values.
    \item \textbf{Validity.} If all honest parties start with the same value $v$, then $v$ is the only value that can be decided by honest parties.
    \item \textbf{No-conflict termination.} If every honest party observes non-conflicting transactions (informally, there exists a unique ``consistent'' value to decide), then every honest party eventually decides.
\end{itemize}
\end{definition}

This relaxation is the core semantic tradeoff: the protocol remains safe under conflicts, but may stop making progress on conflicting state.
